---
title: "P8105_HW2_cc4778"
author: "Chee Kay Cheong"
output: github_document
---

```{r setup, echo = FALSE}
library(tidyverse)

knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

# Problem 1

```{r read clean}
subway_data = read_csv("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

I created a new dataset called '`subway_data`' that consists of 19 variables (columns) and 1868 observations (rows) which were extracted from the original csv datafile '`NYC_Transit_Subway_Entrance_And_Exit_Data.csv`'. So far, I have only cleaned the variable names by making them all lowercase, and I changed the variable type for `entry` from character to logical. 

The 19 variables in this dataset are:

* `line` (character)
* `station_name` (character)
* `station_latitude` (numeric)
* `station_longitude` (numeric)
* `route1` to `route7` (character)  <- 7 variables here
* `route8` to `route11` (numeric)  <- 4 variables here
* `entrance_type` (character)
* `entry` (logical)
* `vending` (character)
* `ada` (logical)

As we can see from the variables and their types, this dataset is not tidy enough. Not all "routes" are character vectors; some routes are numeric. Also, it seems like all the routes should be combined into one variable.

##### Questions

Q1: How many distinct stations are there?
  
A1: **465**
  
```{r stations}
subway_data %>% 
  select(station_name, line) %>% 
  distinct()
```

Q2: How many stations are ADA compliant?

A2: **84**
  
```{r ADA}
subway_data %>%
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct()
```
  
Q3: What proportion of station entrances / exits without vending allow entrance?

A3: **37.7%** of station entrances are not vending allowed. 
  
```{r vending entrance}
subway_data %>% 
  filter(vending == "NO") %>%
  pull(entry) %>% 
  mean
```

*Reformat data so that route number and route name are distinct variables.*
```{r reformat}
subway_data %>% 
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route",
    values_transform = list(route = as.character)) %>% 
  filter(route == "A") %>% 
  select(station_name, line, route) %>% 
  distinct()

subway_data %>%
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route",
    values_transform = list(route = as.character)) %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line, route, ada) %>% 
  distinct()
```
Q4: How many distinct stations serve the A train?

A4: **60** 

Q5: Of the stations that serve the A train, how many are ADA compliant?

A5: **17**




