---
title: "P8105_HW2_cc4778"
author: "Chee Kay Cheong"
output: github_document
---

```{r setup, echo = FALSE}
library(tidyverse)

knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

# Problem 1

```{r read clean}
subway_data = read_csv("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

I created a new dataset called '`subway_data`' that consists of 19 variables (columns) and 1868 observations (rows) which were extracted from the original csv datafile '`NYC_Transit_Subway_Entrance_And_Exit_Data.csv`'. So far, I have only cleaned the variable names by making them all lowercase, and I changed the variable type for `entry` from character to logical. 

The 19 variables in this dataset are:

* `line` (character)
* `station_name` (character)
* `station_latitude` (numeric)
* `station_longitude` (numeric)
* `route1` to `route7` (character)  <- 7 variables here
* `route8` to `route11` (numeric)  <- 4 variables here
* `entrance_type` (character)
* `entry` (logical)
* `vending` (character)
* `ada` (logical)




